Treebeard : An Optimizing Compiler for Tree Based ML Inference

### 2022-04-17 docs/MICRO 2022/probtiling.tex
Optimizations pass

\subsection{Probability Based Tiling}
\subsubsection{More Notation}
In order to formulate the probability based tiling algorithm as an optimization problem, we define the following.
\begin{enumerate}
    \item For every leaf $l \in L$, we define $p_l$ as the probability that the leaf $l$ is reached.
    \item For each node $n \in V$, we define the absolute probability $p_v$ as
    \begin{equation}
        p_v = \begin{cases}
        p_l &\text{if $l \in L$}\\
        p_{left(v)} + p_{right(v)} &\text{otherwise}
        \end{cases}
    \end{equation}
    \item For any tree $T$, $\mathcal{C}(T)$ represents the set of all valid tilings of $T$.
    \item For every $v \in V$, we define $S_v$ as the subtree rooted at $v$.
    \item For every $v \in V$, we define $L_v$ as the set of leaves of $S_v$.
    \item For a every tile $T_i$, we define $root(T_i)$ as the node $v \in T_i$ such that $v$ has no incoming edges from any other node $u \in T_i$.
    \item For a tile $T_i$, $out(T_i) \subseteq E$ is the set of edges $(u, v)$ such that $u \in T_i$ and $v \notin T_i$.
\end{enumerate}

\label{sec:ProbTiling}
\subsubsection{Motivation}
Uniform tiling optimizes tree walks assuming that all root to leaf paths in the tree are equally likely. 
However, we find that this is not always the case. We performed inference on the complete training set of each of our benchmarks and counted how many
times each leaf was reached. While for some benchmarks all leaves are roughly equally likely, for others, a small number of leaves are extremely likely 
while the rest aren't reached often. We call trees with a small number of extremely likely leaves \textbf{\emph{leaf biased}}.
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/airline-ohe.stats.train.txt.png}
    \caption{Statistical profile for airline-ohe}
    \label{Fig:AirlineOHEStats}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/epsilon.stats.train.txt.png}
    \caption{Statistical profile for epsilon}
    \label{Fig:EpsilonStats}
\end{figure}

We say that an input row $r_i$ is \textbf{\emph{covered}} by a subset of leaves $L' \subseteq L$ of a tree $T$, if the leaf $l$ reached by 
walking $T$ for row $r_i$ is in $L'$. We show how different models (and even different trees within the same model) behave differently 
using models for two benchmarks, airline-ohe and epsilon. Consider the graphs shown in   
figures \ref{Fig:AirlineOHEStats} and \ref{Fig:EpsilonStats}. Each line in these graphs corresponds to a fixed fraction of the input (say $f$). 
A point on a line at coordinate $(x, y)$ means that a fraction $y$ of trees in the model could cover a fraction $f$ of all training inputs with a fraction $x$ of 
leaves. For example, the first point on the $f=0.9$ line in figure \ref{Fig:AirlineOHEStats} says that about 52\% of trees ($y$ value) need only 1\% of their
leaves ($x$ value) to cover 90\% of the training input. 
In general, Figure \ref{Fig:AirlineOHEStats} shows that very few leaves are needed to cover a very large fraction of inputs for the benchmark airline-ohe. 
This means that a small fraction of leaves are very likely. On the other hand, for the benchmark epsilon,
figure \ref{Fig:EpsilonStats} shows that a trees need a much larger fraction of their leaves to cover a significant fraction of the test input.
This means that most trees in the epsilon model are not leaf biased.
One other observation we make is that most models have some leaf biased trees while the rest of the trees have equally likely leaves.
\TODO{AP Maybe define a term for trees with roughly equally likely leaves?} We design the probability based tiling algorithm to take advantage of this property 
of decision tree ensembles.

\subsubsection{The Optimization Problem}

@@

We observe that the latency of one tree walk is proportional to the number of tiles that need to be evaluated to reach the leaf.
Therefore, the average inference latency can be modelled as the expected number of tiles that are evaluated to compute one tree prediction. More formally, the problem is to find a tiling $\mathcal{T}$ such that the following objective is minimized.
\[
    \min_{\mathcal{T} \in \mathcal{C}(T)}{\sum_{l \in L} p_l.depth_{\mathcal{T}}(l)}
\]
where the minimization is over all valid tilings $\mathcal{T}$ of the tree $T$, $depth_{\mathcal{T}}(l)$ is the depth of the leaf $l$ given tiling ${\mathcal{T}}$.

The constraints on the optimization are the constraints listed in Section \ref{sec:ValidTiling}. 
The above optimization problem can be solved optimally using dynamic programming. We leave this out in the interest of space. 
Instead, we use the simple greedy algorithm listed in algorithm \ref{Alg:GreedyTilingAlgo} to construct a valid tiling given the node probabilities.
The algorithm starts at the root and greedily keeps adding the most probable legal node to the current tile until the maximum tile size is reached.
Subsequently, the tiling procedure is recursively performed on all nodes that are destinations for edges going out of the constructed tile.

@@

% Talk about problems with increasing number of tile shapes and only performing such tiling on skewed trees
When we tried to apply algorithm \ref{Alg:GreedyTilingAlgo} on all trees in our benchmarks, we found
that even minor variations in probability caused the tiling algorithm to generate a large 
number of tile shapes. This in turn caused a loss in performance because the large size of the 
lookup table needed (section \ref{sec:LookupTable}) caused increased L1 cache misses. In order to 
alleviate this, we only perform probability based tiling on trees that are leaf biased. We decide that 
a tree is leaf biased if a small fraction of leaves, say $\alpha$, can cover a large fraction of training inputs, say $\beta$.
We only perform probability based tiling on skewed trees and fall back to uniform tiling otherwise. 
In our experiments we use $\alpha=0.05$ and $\beta=0.9$.

>>>>

\subsection{Probability Based Tiling}
\label{sec:ProbTiling}
%\subsubsection*{Motivation}

The next algorithm we propose exploits the inherent biases among the leaves of a decision tree. In typical machine learning models some leafs (equivalently outcomes or predictions) are more likely to be reached than others. In such settings, having balanced tiled trees is not sufficient to minimize expected inference time.

Consider for example two machine learning models \op{airline-ohe} and \op{epsilon} (also used in our evaluation). 
Consider the graphs shown in   
figures \ref{Fig:AirlineOHEStats} and \ref{Fig:EpsilonStats} that are generated from training data. Each line in these graphs corresponds to a fixed fraction of the input (say $f$). 
A point on a line at coordinate $(x, y)$ means that a fraction $y$ of trees in the model could cover a fraction $f$ of all training inputs with a fraction $x$ of 
leaves. For example, the first point on the $f=0.9$ line in figure \ref{Fig:AirlineOHEStats} says that about 52\% of trees ($y$ value) need only 1\% of their
leaves ($x$ value) to cover 90\% of the training input. 
In general, Figure \ref{Fig:AirlineOHEStats} shows that very few leaves are needed to cover a very large fraction of inputs for the benchmark \op{airline-ohe}. 
This means that a small fraction of leaves are very likely. 
We call trees with a small number of extremely likely leaves \textbf{\emph{leaf biased}}.

On the other hand, for the benchmark \op{epsilon},
figure \ref{Fig:EpsilonStats} shows that a trees need a much larger fraction of their leaves to cover a significant fraction of the training input.
This means that most trees in the \op{epsilon} model are not leaf biased.

\CommentOut{
\subsubsection{More Notation}
In order to formulate the probability based tiling algorithm as an optimization problem, we define the following.
\begin{enumerate}
    \item For every leaf $l \in L$, we define $p_l$ as the probability that the leaf $l$ is reached.
    \item For each node $n \in V$, we define the absolute probability $p_v$ as
    \begin{equation}
        p_v = \begin{cases}
        p_l &\text{if $l \in L$}\\
        p_{left(v)} + p_{right(v)} &\text{otherwise}
        \end{cases}
    \end{equation}
    \item For any tree $T$, $\mathcal{C}(T)$ represents the set of all valid tilings of $T$.
    \item For every $v \in V$, we define $S_v$ as the subtree rooted at $v$.
    \item For every $v \in V$, we define $L_v$ as the set of leaves of $S_v$.
    \item For a every tile $T_i$, we define $root(T_i)$ as the node $v \in T_i$ such that $v$ has no incoming edges from any other node $u \in T_i$.
    \item For a tile $T_i$, $out(T_i) \subseteq E$ is the set of edges $(u, v)$ such that $u \in T_i$ and $v \notin T_i$.
\end{enumerate}
}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/airline-ohe.stats.train.txt.png}
    \caption{Statistical profile for airline-ohe}
    \label{Fig:AirlineOHEStats}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/epsilon.stats.train.txt.png}
    \caption{Statistical profile for epsilon}
    \label{Fig:EpsilonStats}
\end{figure}
\CommentOut{
We say that an input row $r_i$ is \textbf{\emph{covered}} by a subset of leaves $L' \subseteq L$ of a tree $T$, if the leaf $l$ reached by 
walking $T$ for row $r_i$ is in $L'$. We show how different models (and even different trees within the same model) behave differently 
using models for two benchmarks, airline-ohe and epsilon. Consider the graphs shown in   
figures \ref{Fig:AirlineOHEStats} and \ref{Fig:EpsilonStats}. Each line in these graphs corresponds to a fixed fraction of the input (say $f$). 
A point on a line at coordinate $(x, y)$ means that a fraction $y$ of trees in the model could cover a fraction $f$ of all training inputs with a fraction $x$ of 
leaves. For example, the first point on the $f=0.9$ line in figure \ref{Fig:AirlineOHEStats} says that about 52\% of trees ($y$ value) need only 1\% of their
leaves ($x$ value) to cover 90\% of the training input. 
In general, Figure \ref{Fig:AirlineOHEStats} shows that very few leaves are needed to cover a very large fraction of inputs for the benchmark airline-ohe. 
This means that a small fraction of leaves are very likely. On the other hand, for the benchmark epsilon,
figure \ref{Fig:EpsilonStats} shows that a trees need a much larger fraction of their leaves to cover a significant fraction of the test input.
This means that most trees in the epsilon model are not leaf biased.
One other observation we make is that most models have some leaf biased trees while the rest of the trees have equally likely leaves.
\TODO{AP Maybe define a term for trees with roughly equally likely leaves?} We design the probability based tiling algorithm to take advantage of this property 
of decision tree ensembles. 
}
\subsubsection{The Optimization Problem}

@@

Observe that the latency of one tree walk is proportional to the number of tiles that need to be evaluated to reach the leaf. It is easy to see that for a leaf biased tree, basic tilling does not optimize for this objective, it considers all leafs to be equally likely.

The goal of probablistic tiling is to minimize the average inference latency, or equivalently the minimize the expected number of tiles that are evaluated to compute one tree prediction. More formally, the problem is to find a \emph{valid} (as defined in Section~\ref{sec:ValidTiling}) tiling $\mathcal{T}$ such that the following objective is minimized.
\[
    \min_{\mathcal{T} \in \mathcal{C}(T)}{\sum_{l \in L} p_l.depth_{\mathcal{T}}(l)}
\]
where the minimization is over all valid tilings $\mathcal{T}$ of the tree $T$, $depth_{\mathcal{T}}(l)$ is the depth of the leaf $l$ given tiling ${\mathcal{T}}$. $p_l$ is the probability of of reaching leaf $l$ as observed during training.

The above optimization problem can be solved optimally using dynamic programming. 
We leave this out in the interest of space. 
Instead, we use the simple greedy algorithm listed in algorithm \ref{Alg:GreedyTilingAlgo} to construct a valid tiling given the node probabilities\footnote{Probabilites for internal nodes can be computed from probablities for leafs by summing up the probabilities of all leafs that belong to the sub-tree rooted at the internal node. Leaf probabilities are collected during training.}.
The algorithm starts at the root and greedily keeps adding the most probable legal node to the current tile until the maximum tile size is reached.
Subsequently, the tiling procedure is recursively performed on all nodes that are destinations for edges going out of the constructed tile.

@@

% Talk about problems with increasing number of tile shapes and only performing such tiling on skewed trees
\CommentOut{
When we tried to apply algorithm \ref{Alg:GreedyTilingAlgo} on all trees in our benchmarks, we found
that even minor variations in probability caused the tiling algorithm to generate a large 
number of tile shapes. This in turn caused a loss in performance because the large size of the 
lookup table needed (section \ref{sec:LookupTable}) caused increased L1 cache misses. In order to 
alleviate this, we only perform probability based tiling on trees that are leaf biased.
}
We find probability based tiling is only beneficial for leaf biased trees\footnote{Turns out that probability based tiling produces many more tile shapes (see Section~\ref{sec:tileShapes}) which direclty impacts the cost of \op{getChildTile} making it more expensive than basic tiling}.  Recall that   
a tree to be leaf biased if a small fraction of leaves, say $\alpha$, can cover a large fraction of training inputs, say $\beta$.
We only perform probability based tiling on trees with thresholds $\alpha=0.05$ and $\beta=0.9$ and fall back to uniform tiling otherwise.

### 2022-04-17 docs/MICRO 2022/uniftiling.tex
Optimizations pass

\subsubsection{Further Opimization and Code Generation}
We found that most leaf tiles for a given tree are at the same depth when uniform tiling is used. Furthermore, we see that deeper leaves 
are more likely to be reached.
%\footnote{Intuitively, this is true because training algorithms keep splitting nodes to maximize gain and gain
%will typically be maximized by splitting a large number of inputs.}.
Based on these observations, we (optionally) pad the tree of tiles generated with uniform tiling so that all leaves are at the same depth.
This transformation is performed on the high level IR after uniform tiling. 
Once the trees have been padded to make all leaves equal depth, the tree walks are fully unrolled to evaluate a fixed 
number of tiles and all leaf checks are omitted.

@@

\TODO{AP: This listing is unnecesarily long. Can we maybe leave out the loop bodies and say something like "depth 2 walk"? Should 
we point to the figure in the overview section instead?}

>>>>

One interesting property of this tilling algorithm is that it naturally reduces the imbalance in trees, especially at large tile sizes. As the algorithm traverses down to sparser level of the tree, it naturally groups sub-trees containing chains of nodes, thus balancing the trees. While its possible to further enhance the algorithm to explicitly balance tiled trees, we find that basic tiling suffices in practice.
\CommentOut{
\subsubsection{Further Opimization and Code Generation}
We found that most leaf tiles for a given tree are at the same depth when uniform tiling is used. Furthermore, we see that deeper leaves 
are more likely to be reached.
%\footnote{Intuitively, this is true because training algorithms keep splitting nodes to maximize gain and gain
%will typically be maximized by splitting a large number of inputs.}.
Based on these observations, we (optionally) pad the tree of tiles generated with uniform tiling so that all leaves are at the same depth.
This transformation is performed on the high level IR after uniform tiling. 
Once the trees have been padded to make all leaves equal depth, the tree walks are fully unrolled to evaluate a fixed 
number of tiles and all leaf checks are omitted.

@@

\TODO{AP: This listing is unnecesarily long. Can we maybe leave out the loop bodies and say something like "depth 2 walk"? Should 
we point to the figure in the overview section instead?}
}
